---
title: "Tidytuesday Malaria"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggmap)
library(countrycode)
library(grid)
library(gridExtra)
```

```{r, message=FALSE}
malaria <- read_csv("malaria_deaths.csv") %>%
  setNames(c("country", "code", "year", "deaths"))
```

How many countries are in the dataset? 

```{r}
malaria %>%
  count(country)
```

Ok, we have around 27 datapoints for each country. 

Let's see see which countries are especially plagued by malaria:

```{r}
(plagued_countries <- malaria %>%
  group_by(country) %>%
  summarise(
    deaths = sum(deaths)
  ) %>%
  arrange(desc(deaths)) %>%
  head)
```

Ok, Sierra Leone seems to be especially affected. Others are Burkina Faso, Uganda, Equatorial Guinea, Cote d'Ivoire, and Nigeria. 

Let's plot the deaths for each country overall:

```{r}
world <- map_data("world")

malaria_some_years <- malaria %>%
  filter(year %in% c(1995, 2000, 2005, 2010))

deaths_per_year <- malaria_some_years %>%
  inner_join(iso3166 %>% select(a3, mapname), by = c(code = "a3")) %>%
  left_join(world, by = c(country = "region"))

ggplot(deaths_per_year, aes(long, lat, map_id = mapname, 
                                                fill = deaths)) +
  geom_map(map = world) +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_void() +
  coord_map(xlim = c(-180, 180)) +
  labs(fill = "Deaths per year") +
  facet_wrap(~ year, ncol = 2)
```

Let's only display Africa

```{r}
deaths_africa_per_year <- deaths_per_year %>%
  mutate(
    countrycode(deaths_per_year$mapname, 
                origin = "country.name", 
                destination = "continent")) %>%
  rename(continent = `countrycode(...)`) %>%
  filter(continent == "Africa")

ggplot(deaths_africa_per_year, aes(long, lat, map_id = mapname, 
                                   fill = deaths)) +
  geom_map(map = world) +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_void() +
  coord_map(xlim = c(-80, 80)) +
  labs(fill = "Deaths per year") +
  facet_wrap(~ year, ncol = 2)
```

How did the deaths of malaria change globally since 1990? 

```{r}
malaria %>%
  group_by(year) %>%
  summarise(deaths = sum(deaths)) %>%
  ggplot(aes(x = year, y = deaths)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 4500)) +
  scale_x_continuous(breaks = seq(1990, 2016, by = 5)) +
  labs(title = "Malaria deaths per year worldwide",
       y = "Deaths",
       x = "Year")
```

Which countries actually improved the most over the past decades?

```{r}
malaria %>%
  group_by(year, country) %>%
  summarise(deaths = sum(deaths)) %>%
  arrange(country, year) %>%
  filter(country %in% plagued_countries$country) %>%
  ungroup() %>%
  mutate(
    improvement = deaths - lag(deaths)
  ) %>%
  ggplot(aes(x = year, y = improvement, colour = country)) +
  geom_line() +
  ylab("Improvements (lower = better)")
```


